A Theory for Migration of Information Systems

### 2022-08-28 2022 Migration/articleMigration.tex
Further refinement on the definition of datasets

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\id#1{\text{\it #1\/}}
\def\Events{{\mathit E}}
\begin{document}

@@

>>>>

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\Events{{\mathit E}}
\begin{document}
\include{preamramics}

@@

\section{Basics}
\label{sct:Basics}
	In this section we explain the basic notions to describe migration.
	We introduce the notion of dataset.
	A dataset is a quintuple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
	which consists of a set of rules $\rules$, a set of relations $\rels$, a set of concepts $\concepts$, a set of pairs $\pairs$, and a set of atoms $\atoms$.
	Let us introduce these five notions one by one.
\subsection{Atoms}
	Atoms (from set $\atoms$) are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.
    Each atom is an instance of a \emph{concept}.
\subsection{Concepts}
	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``A woman is a human.'' or ``A human is a mammal.''.
\subsection{Pairs}
	Pairs (from set $\pairs$) are used to relate two atoms.
	A pair $\pair{\text{\atom{Peter}}}{\text{\atom{074238991}}}$ might bear the meaning that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	In the formal world, we do not care about that meaning,
	leaving it entirely up to a user which meaning (s)he attaches to this pair.
\subsection{Relations}
	Relations (from set $\rels$) are used in information systems to store data as pairs.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	Every relation $r$ has a set of pairs, $\pop{r}$, which we call the population of $r$.
	Every relation has a source concept and a target concept.
	If we need to be specific about the source and target concepts, we write $\declare{r}{A}{B}$ instead of just $r$.
	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.

	The signature rule guarantees that each pair in $r$ respects the signature of $r$:
\begin{equation}
	\tag{signature rule}
	\pair{a}{b}\in\pop{\declare{r}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}

	The population rule defines the population of a dataset:
\begin{equation}
	\tag{population rule}
	\pairs\ =\ \{ \pair{a}{b}|\ r\in\rels,\ \pair{a}{b}\in\pop{r}\} \label{def. population}
\end{equation}

	The atoms rule defines all atoms in a dataset:
\begin{equation}
	\tag{atoms rule}
	\atoms\ =\ \{ a |\ \pair{b}{c}\in\pairs,\ a\in\{b,c\}\ \} \label{def atoms}
\end{equation}
\subsection{Terms}
	A term is a combination of relations (from $\rels$), which we use to express constraints from the real world.
	For this purpose we use the operators from relation algebra~\cite{BrinkKahlSchmidt1997}.
\begin{definition}[terms]
\label{def:terms}
\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
\begin{eqnarray*}
	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
\end{eqnarray*}
\end{definition}
	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.

	The relations used in a term $t$ are given by $\relsIn{t}$:
\begin{definition}[relations in terms]
	\label{relations in terms}
	\item   For every $A,B\in\concepts$ and $r,s\in\terms$
	\begin{eqnarray*}
		\relsIn{r}		 &=&\{r\}	\\
		\relsIn{r \cup s}	 &=&\relsIn{r}\cup\relsIn{s}\\
		\relsIn{r \cap s}	 &=&\relsIn{r}\cup\relsIn{s}\\
		\relsIn{r-s}	 &=&\relsIn{r}\cup\relsIn{s}\\
		\relsIn{r;s}	 &=&\relsIn{r}\cup\relsIn{s}\\
		\relsIn{r\backslash s}	 &=&\relsIn{r}\cup\relsIn{s}\\
		\relsIn{r\slash s}	 &=&\relsIn{r}\cup\relsIn{s}\\
		\relsIn{\flip{r}}	 &=&\relsIn{r}\\
		\relsIn{\cmpl{\declare{r}{A}{B}}}	 &=&\relsIn{r}\\
	%	\relsIn{\kleeneplus{r}}	 &=&\relsIn{r}\\
	%	\relsIn{\kleenestar{r}}	 &=&\relsIn{r}\\
		\relsIn{\ident{A}} 	 &=&\emptyset\\
		\relsIn{\fullt{A}{B}}&=&\emptyset\\
	\end{eqnarray*}
\end{definition}
	The \define{signature} of a term $r$ is a pair of concepts, given by $\tf{r}$.
	$\mathfrak{T}$ is a function that maps terms to signatures.
	This function and its peculiarities are discussed in~\cite{Joosten2015}
	and have no consequences for the remainder of this paper.
	(\emph{\@Bas, komen we ermee weg als we de definitie van $\tf{r}$ hier weglaten?})

	Every term $t$ has a set of pairs $\pop{t}$, which we call the population of that term.
 	Let $A$ and $B$ be finite sets of atoms, then the population of term is defined as follows:
\begin{definition}[population of terms]
\label{population of terms}
\item   For every $A,B\in\concepts$ and $r,s\in\terms$
\begin{eqnarray*}
	\pop{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
	\pop{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\pop{r}\ \text{or }\ \pair{a}{b}\in\pop{s}\}	\\
	\pop{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\pop{r}\ \text{and}\ \pair{a}{b}\in\pop{s}\}	\\
	\pop{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\pop{r}\ \text{and}\ \pair{a}{b}\notin\pop{s}\}	\\
	\pop{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\pop{r}\ \text{and}\ \pair{b}{c}\in\pop{s}\}	\\
	\pop{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\pop{r}\ \text{implies}\ \pair{a}{c}\in\pop{s}\}	\\
	\pop{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\pop{s}\ \text{implies}\ \pair{a}{c}\in\pop{r}\}	\\
	\pop{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\pop{r}    \}	\\
	\pop{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
%	\pop{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
%	\pop{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
	\pop{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
	\pop{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
\end{eqnarray*}
\end{definition}

\subsection{Rules}
	A rule (from set $\rules$) is an equality between two terms that constrains the data stored in relations,
	to ensure that all rules in the dataset are satisfied.
	A \define{rule} is a pair of terms $r,s\in\terms$ with the same types ($\tf{r}=\tf{s}$), which we denote as:
\[\text{RULE}\ r = s\]
	This means that the constraint \(\pop{r} = \pop{s}\) is satisfied throughout the dataset.
	In practice, we will also use \(\text{RULE}\ r\subs s\) as a shorthand for \(\text{RULE}\ r\cap s = r\).

	We call a rule $\text{RULE}\ r = s$ \define{satisfied} if \(\pop{r} = \pop{s}\).
	Each relation that is used in a term of a dataset $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ is an element of $\rels$ as well:
\begin{equation}
	\tag{relations in dataset}
	\{ r|\ \text{RULE}\ t = s\in\rules,\ r\in\relsIn{t}\cup\relsIn{s}\ \}\subseteq\rels\label{def relations in dataset}
\end{equation}

	The data in a data set $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ are the pairs in the population of all relations in $\rels$.
	This constitutes the set $\pairs$, also known as the population of the dataset.
	In response to messages such as
\[\text{INSERT}\ <3,{\tt Brown}> \text{INTO}\ \declare{r}{Age}{Name}\]
 	the population of a dataset changes. 
	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
	We call a rule \define{violated} if it is not satisfied.
	The set $\pop{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
	Consequently, such a rule may also be referred to as \define{invariant}.
	Each pair in the violation set of a rule is called a violation of that rule.
